Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-16 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Ch  8 images

The goal of this chapter is to utilize an analytical tool to predict or simulate the results from some of the key experiments conducted in this study. Unlike the previous chapter the computational fluid dynamics (CFD) models are not as widely used by fire investigators as the algebraic tools are.  The CFD models require more knowledge, more computer resources, and more input data to utilize correctly.  Currently modelers need to develop their own data or rely on poorly defined databases with the hope that the material tested in the database is similar to the specific material of interest.

@@

In this section, the natural gas burner will be modelled with a range of grid cell resolutions and radiative fraction values.  The results are compared against the source fire characterization data to determine which grid cell resolution and radiative fraction will provide the best fire simulation for use in the simulations of the wall and compartment experiments.

@@

To examime the impact of computational grid cell size on the results, three different computational cell sizes; 80~mm, 40~mm, and 20~mm were choosen to model the fire. The resulting mesh resolution index (RI) values were approximately 4.4, 8.8, and 17.5, respectively. Given the range of grid cell sizes and the characteristic diameter of fire, 0.35~m, this resulted in what could be considered a course (RI=5), medium (RI=10), and fine (RI=20) mesh resolution.  FDS simulations of the source fire characterization experiments will be conducted with these mesh resolutions and compared with the results from the experiments.

@@

The burner was modeled as an obstruction with a open top area of 0.09 m$^2$, elevated 0.08~m above the test surface.  This is the same physical size of the natural gas burner used in the experiments.  It was modeled with a steady state heat release rate flux of 888.89 kW/m$^2$, or a heat release rate of 80~kW.   The chosen fuel reaction was `methane' since the natural gas used in the laboratory experiments contained approximately 95~\% methane.  In FDS version 6, the default radiant fraction for the `methane' reaction was set to 0.2.

FDS has devices, which are used to simulate thermocouples and total heat flux gauges.  These devices were positioned in locations similar to the locations used in the source fire experiments.  The thermocouples were positioned in a vertical array, centered on the burner, beginning 0.1~m above the burner surface and spaced every 0.1~m upto 1.2~m above the burner surface.  A total heat flux gauge was also modelled to provide data at the 0.95 m from the centerline of the burner and 0.5 m above the top of the burner.  
In addition, FDS features devices that can not be physically implemented.  One such device, is used to determine the heat release rate per unit length.  This device was used to determine flame height based on that measure.

@@

\subsection{Grid Size Analysis}
Figure~\ref{FHNG01_RI} shows the comparison of the centerline gas burner temperatures with three different sets of FDS results.  The only variable in the simulations is the RI. The RIs include three different computational cell sizes; 80~mm (RI=5), 40~mm~(RI=10), 20~mm~(RI=20).  The associated 95~\% confidence limits are also shown the temperature value for each height.  The FDS simulated temperatures based on the course cell size of 80~mm (RI=5) had the best agreement with the temperatures along centerline of the plume.  The two simulations with the higher RI values provided less agreement than the course mesh.

@@

This seemed to be counter intuitive to the results expected from an increased RI.  However in each of the FDS models, the results closest to the burner seem to indicate that the model may not be `burning' the appropriate amount of fuel (fuel rich), which resulted in temperatures significantly lower than measured.  The effect of this fuel rich area above the burner results in poorer agreement with increasing RI.  It appears that the the RI=10 and RI=20 models follow the correct trend, however the combuston zone was moved to a higher elevation relative to what was measured.  For example the values at 0.5~m and 0.8~m above the burner, from the RI=10 and RI=20 cases respectively, seemed to coincide with the measured value at 0.3~m above the burner.  Making the adjustment appears to move the remaining temperatures above those points into alignment with the measurements along the centerline of the plume. As FDS is a complex model, this explaination may be only one possibility.

These simulation results are consistant with the comparisons made to the McCaffery's plume correlation from his study, `Purely Buoyant Diffusion Flames'~\cite{McCaffrey:1979}, in the FDS Validation Guide~\cite{FDS_Validation_Guide}. McCaffery used a natural gas burner, 0.30~m square, which is similar in size to the burner used in this study.  McCaffery noted that, "visually non-axisymmetric behavior could be observed only very close to the burner surface".  In the study, temperatures were measured with Type K thermocouples covered with a 1.0~mm stainless steel sheath. The peak heat release rate in McCaffery's plume study was 57.5~kW, this heat release rate was also the closest to the 80~kW fires in the current study.  The comparison between the McCaffery plume correlation and the FDS simulations at RI values of 5, 10 and 20 are shown in Figure~\ref{McCaffrey_Plume_comp}. Focusing on the data between the burner surface and the 1~m above the burner surface, FDS underpredicted the temperature near the surface of the burner and then overpredicted as the elevation increased to approximately 0.4~m and 1.3~m above the burner for the RI=10 and 20 cases respectively. Once the temperature comparisons were above the continuous and intermittant flame regions, the agreement between the cases and the correlation began to improve.  Similar to the current study, increased RIs did not improve the simulation of the vertical centerline temperatures close to the burner.

@@

The comparisons above used the predicted gas temperature from the model. Another comparison can be made examining the `thermocouple' results from FDS to see if the model of the thermocouple provides improved argeement with the temperatures measured via physical thermocouples.  This comparison is shown in Figure~\ref{FHNG01_RI_TC}.  Not surprisingly, the sub model of the thermocouple device shifted the FDS values slightly.  At temperatures greater than 700~$^\circ$C, the `thermocouple' temperatures tended to be lower than the gas temperatures, perhaps due to radiant heat transfer losses. At temperatures less than 700~$^\circ$C, the `thermocouple' temperatures tended to be a higher than the gas temperatures, perhaps due to gains from radiant heat transfer.  The best agreement along the vertical array of temperatures was provided by the RI=5 model.  Variables such as density and emissivity required for modeling the thermal response and heat transfer to the thermocouple bead are generating the diffrences between the gas temperature and the measured temperature FDS values.

@@

From the Chapter 2 source fire characterization experiments, the measured heat flux at 0.95~m from the centerline of the burner and 0.5~m above the top of the burner. 
In Figure~\ref{FHNG01_HF_RI} the averaged values from both heat flux gages are plotted against the FDS calculated heat flux at three different RIs of 5, 10 and 20.  These FDS simulations used the default radiative fraction value of 0.2.  As with the temperature comparisons, the course grid with an RI equal to 5 provided the best agreement to the measurements, however all of the values were lower than the measured heat flux.

@@

One last check on the plume temperature predictions was made to examine the impact of the range of radiative fractions on the FDS results.  The comparisons were made using the course grid, RI=5, the radiative fraction values of 0.20, 0.22, 0.25, and 0.30, as before. The simulations with radiative fractions between 0.2 and 0.25 were within the 2~$\sigma$ limits of the temperature measurements at each elevation.

@@

One of the outputs that can be gotten from FDS is a heat release rate per unit length (HRRPUL), it has been used to determine the height when energy stops being released along the vertical centerline of the gas burner.  Estimates of the flame height were made by estimating the height where 0.95 to 0.99~$\%$ of the total energy had been released. Table~\ref{tab:FDS_FH} displays estimated values from the models using all three RI values as well as the 95~$\%$ certainty range for the peak visible flame height from Chapter 2.  The course and medium grids, RI=5 and R=10, are not significantly different from the estimations of the peak visible flame height 95~$\%$ certainty value range.

\begin{table}[h]
  \small
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline Peak Visible Flame Height   	& RI=5       		& RI=10      		& RI=20  \\
    Range          						&  (80~mm grid)  	&  (40~mm grid)  	&  (20~mm grid)  \\
    (m)     							& (m) 				& (m) 				& (m)  			 \\ \hline
    \hline 0.73 to 1.14         		& 1.02 to 1.27      & 1.07 to 1.27      & 1.28 to 1.42   \\
    \hline
  \end{tabular}
  \caption[Comparison of estimations of flame height from the natural gas burner experiments and FDS simulations with RIs of 5, 10 and 20.]{Comparison of estimations of flame height from the natural gas burner experiments and FDS simulations with RIs of 5, 10 and 20.}
  \label{tab:FDS_FH}
\end{table}

@@

The comparisons of the temperature, heat flux, and flame heights for this specific scenerio show the best agreement with the 80~mm grid (RI=5) FDS simulations.  Even though the differences between the radiative fractions of 0.2 and 0.22 are small, for the remaining simulation cases, 0.22 will be used.  While the best agreement for the fire was demonstrated with the RI= 5 grid, the RI=10 grid will also be used in the subsequent instrumented wall and the compartment model sections to see if the increased resolution comes into play as the models get additional obstructions and surfaces.

@@

The next comparsions will be made with the natural gas source fire at 4 different distances from an instrumented gypsum board wall.  Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610~m), one burner width (0.305~m), one half burner width (0.150~m), and then touching the wall (0.0~m).  The simulations will replicate the same distances.  The simulation includes a vertical gypsum board wall, 1.2~m wide, that extends from the floor of the model up to the top of the domain.  The material properties of the wall are based solely on the properties of gypsum that were cited in Chapter 4.  The values that are compared in the section include: vertical centerline temperatures above the burner, wall temperatures, and heat flux values at the position of the wall surface.

@@

As before, the FDS simultions will include devices which are used to simulate thermocouples and total heat flux gauges.  These devices were positioned in locations similar to the locations used in the physical experiments, above the centerline of the burner and on the wall as shown in Chapter 4, Figure~\ref{Instrumented_GB_Wall}. In addition, HRRPUL will again be used as a means to gain insight into the peak flame height simluated by FDS.

@@

 There seem to be a couple of items that stick out from the graphs.  First is what appears to be a separation of the temperature data into three distinct sections.  The bottom three positions, 0.2~m through 0.6~m, represent the hottest section.  The next two measurement positions at 0.8~m and 1.0~m are similar in temperature.  The third section extends from 1.2 through 2.2~m above the burner and has the shape of a smooth curve decreasing in temperature with increasing elevation.  Comparing the temperatures to the flame height analysis in Chapter 2, the first section aligns with the median flame height, the second section aligns with the portion of the flame that is intermittent, 40~$\%$ of the time or less.  The upper most section of the temperature data aligns with the thermal plume above the flame.

 The second issue is that the FDS predictions do not align equally well with the data from each position, with the best agreement occurring when the burner is positioned against the wall.  Further the temperatures for the burner position, 0.61~m from the was thought to be equivalent to the source fire experiments documented in Chapter 2. However that is not the case.  The installation of the wall resulted in movement of the flame and plume.  The flames oscillate toward the wall and as a result move away from the thermocouples centered over the burner. The resulting decrease in temperature may account for the difference between the plume tempertaures and the simulated temperatures from FDS.

@@

@@

Complete to compare measured flame height with FDS cal~\cite{McDermott:2011}

>>>>

The goal of this chapter is to utilize an analytical tool to predict or simulate the results from some of the key experiments conducted in this study.  Computational fluid dynamics (CFD) models are not as widely used by fire investigators as the algebraic tools are.  The CFD models require more knowledge, more computer resources, and more input data to utilize correctly.  Currently modelers need to develop their own data or rely on loosely defined databases with the hope that the material tested in the database is similar to the specific material of interest.

@@

In this section, the natural gas burner will be modelled with a range of grid cell resolutions and radiative fraction values.  The results are compared against the source fire characterization data to determine which grid cell resolution and radiative fraction provided the best fire simulation for use in the simulations of the wall and compartment experiments.

@@

To examime the impact of computational grid cell size on the results, three different computational cell sizes; 80~mm, 40~mm, and 20~mm were choosen to model the fire. The resulting resolution index (RI) values were approximately 4.4, 8.8, and 17.5, respectively. Given the range of grid cell sizes and the characteristic diameter of fire, 0.35~m, this resulted in what could be considered a course (RI=5), medium (RI=10), and fine (RI=20) resolution.  FDS simulations of the source fire characterization experiments will be conducted with these resolutions and compared with the results from the experiments.

@@

The burner was modeled as an obstruction with a open top area of 0.09~m$^2$, elevated 0.08~m above the test surface.  This is the same physical size of the natural gas burner used in the experiments.  It was modeled with a steady state heat release rate flux of 888.89 kW/m$^2$, or a heat release rate of 80~kW.   The chosen fuel reaction was `methane' since the natural gas used in the laboratory experiments contained approximately 95~\% methane.  In FDS version 6, the default radiant fraction for the `methane' reaction was set to~0.2.

FDS has devices, which are used to simulate thermocouples and total heat flux gauges.  These devices were positioned in locations similar to the locations used in the source fire experiments.  The thermocouples were positioned in a vertical array, centered on the burner, beginning 0.1~m above the burner surface and spaced every 0.1~m upto 1.2~m above the burner surface.  A total heat flux gauge was also modelled to provide data at the 0.95 m from the centerline of the burner and 0.5 m above the top of the burner.  
In addition, FDS features devices that can not be physically implemented.  One such device, is used to determine the heat release rate per unit length.  This device was used to determine flame height~\cite{McDermott:2011}.

@@

\subsection{Grid Size Analysis}
Figure~\ref{FHNG01_RI} shows the comparison of the centerline gas burner temperatures with three different sets of FDS results.  The only variable between these simulations was the RI. The RIs included three different computational cell sizes; 80~mm (RI=5), 40~mm~(RI=10), 20~mm~(RI=20).  The associated 95~\% confidence limits are also shown the temperature value for each height.  The FDS simulated temperatures based on the course cell size of 80~mm (RI=5) had the best agreement with the temperatures along centerline of the plume.  The two simulations with the higher RI values provided less agreement than the course mesh.

@@

This seemed to be counter intuitive to the results expected from an increased RI.  However in each of the FDS models, the results closest to the burner seem to indicate that the model may not be `burning' the appropriate amount of fuel (fuel rich), which resulted in temperatures significantly lower than measured.  The effect of this fuel rich area above the burner results in poorer agreement with increasing RI.  It appears that the the RI=10 and RI=20 models follow the correct trend, however the combustion zone was moved to a higher elevation relative to what was measured.  For example the values at 0.5~m and 0.8~m above the burner, from the RI=10 and RI=20 cases respectively, seemed to coincide with the measured value at 0.3~m above the burner.  Adjusting the FDS corresponding predictions from the RI=10 and RI=20 models down to the 0.3~m position appears to move the remaining temperatures above those points into alignment with the measurements along the centerline of the plume. As FDS is a complex model, this explaination may be only one possibility.

These simulation results are consistant with the comparisons made to the McCaffrey's plume correlation from his study, `Purely Buoyant Diffusion Flames'~\cite{McCaffrey:1979}, in the FDS Validation Guide~\cite{FDS_Validation_Guide}. McCaffrey used a natural gas burner, 0.30~m square, which is similar in size to the burner used in this study.  McCaffrey noted that, "visually non-axisymmetric behavior could be observed only very close to the burner surface".  In the McCaffrey study, temperatures were measured with Type K thermocouples covered with a 1.0~mm stainless steel sheath. The peak heat release rate in McCaffery's plume study was 57.5~kW, this heat release rate was also the closest to the 80~kW fires in the current study.  The comparison between the McCaffrey plume correlation and the FDS simulations at RI values of 5, 10 and 20 are shown in Figure~\ref{McCaffrey_Plume_comp}. Focusing on the data between the burner surface and the 1~m above the burner surface, FDS underpredicted the temperature near the surface of the burner and then overpredicted as the elevation increased to approximately 0.4~m and 1.3~m above the burner for the RI=10 and 20 cases respectively. Once the temperature comparisons were above the continuous and intermittant flame regions, the agreement between the cases and the correlation began to improve.  Similar to the current study, increased RIs did not improve the simulation of the vertical centerline temperatures close to the burner.

@@

The comparisons above used the predicted gas temperature from the model. Another comparison can be made examining the `thermocouple' results from FDS to see if the model of the thermocouple provides improved argeement with the temperatures measured via physical thermocouples.  This comparison is shown in Figure~\ref{FHNG01_RI_TC}.  Not surprisingly, the sub model of the thermocouple device shifted the FDS values slightly.  At temperatures greater than 700~$^\circ$C, the `thermocouple' temperatures tended to be lower than the gas temperatures, perhaps due to radiant heat transfer losses. At temperatures less than 700~$^\circ$C, the `thermocouple' temperatures tended to be a higher than the gas temperatures, perhaps due to gains from radiant heat transfer.  The best agreement along the vertical array of temperatures was provided by the RI=5 model.  Variables such as density and emissivity required for modeling the thermal response and heat transfer to the thermocouple bead may be generating the differences between the gas temperature and the measured temperature FDS values.

@@

The heat flux data being used for comparison was from the Chapter 2 source fire characterization experiments, the measured heat flux at 0.95~m from the centerline of the burner and 0.5~m above the top of the burner.  In Figure~\ref{FHNG01_HF_RI} the averaged values from both heat flux gages are plotted against the FDS calculated heat flux at three different RIs of 5, 10 and 20.  These FDS simulations used the default radiative fraction value of~0.2.  As with the temperature comparisons, the course grid with an RI=5 provided the best agreement to the measurements, however all of the values were lower than the measured heat flux.

@@

Another examinination of the plume temperature predictions was made to examine the impact of the range of radiative fractions on the FDS results.  The comparisons were made using the course grid, RI=5, and radiative fraction values of 0.20, 0.22, 0.25, and 0.30, as before. The simulations with radiative fractions between 0.2 and 0.25 were within the 2~$\sigma$ limits of the temperature measurements at each elevation.

@@

One of the outputs that can be gotten from FDS is a heat release rate per unit length (HRRPUL), it has been used to determine the height when energy stops being released along the vertical centerline of the gas burner~\cite{McDermott:2011}.  Estimates of the flame height were made by estimating the height where 0.95 to 0.99~$\%$ of the total energy had been released. Table~\ref{tab:FDS_FH} displays estimated values from the models using all three RI values as well as the 95~$\%$ certainty range for the peak visible flame height from Chapter 2.  The course and medium grids, RI=5 and R=10, were not significantly different from the estimations of the peak visible flame height 95~$\%$ certainty value range.

\begin{table}[h]
  \small
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline Peak Visible Flame Height   	& RI=5       		& RI=10      		& RI=20  \\
    Range          						&  (80~mm grid)  	&  (40~mm grid)  	&  (20~mm grid)  \\
    (m)     							& (m) 				& (m) 				& (m)  			 \\ 
    \hline 0.73 to 1.14         		& 1.02 to 1.27      & 1.07 to 1.27      & 1.28 to 1.42   \\
    \hline
  \end{tabular}
  \caption[Comparison of estimations of flame height from the natural gas burner experiments and FDS simulations with RIs of 5, 10 and 20.]{Comparison of estimations of flame height from the natural gas burner experiments and FDS simulations with RIs of 5, 10 and 20.}
  \label{tab:FDS_FH}
\end{table}

@@

The comparisons of the temperature, heat flux, and flame heights for this specific scenerio showed the best agreement with the 80~mm grid (RI=5) FDS simulations.  Even though the differences between the radiative fractions of 0.2 and 0.22 were small, for the remaining simulation cases, 0.22 will be used.  While the best agreement for the fire was demonstrated with the RI= 5 grid, the RI=10 grid will also be used in the subsequent instrumented wall and the compartment model sections to see if the increased resolution comes into play as the models get additional obstructions and surfaces.

@@

The next comparsions were made with the natural gas source fire at 4 different distances from an instrumented gypsum board wall.  Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610~m), one burner width (0.305~m), one half burner width (0.150~m), and then touching the wall (0.0~m).  The simulations will replicate the same distances.  The simulation included a vertical gypsum board wall, 1.2~m wide, that extends from the floor of the model up to the top of the domain.  The material properties of the wall are based solely on the properties of gypsum that were cited in Chapter 4.  The values that are compared in the section include: vertical centerline temperatures above the burner, wall temperatures, and heat flux values at the position of the wall surface.

@@

As before, the FDS simulations included devices which are used to simulate thermocouples and total heat flux gauges.  These devices were positioned in locations similar to the locations used in the physical experiments, above the centerline of the burner and on the wall as shown in Chapter 4, Figure~\ref{Instrumented_GB_Wall}. In addition, HRRPUL will again be used as a means to gain insight into the peak flame height simulated by FDS.

@@

 There seem to be a couple of items that stick out from the graphs.  First, there appears to be a separation of the temperature data into three distinct sections.  The bottom three positions, 0.2~m through 0.6~m, represent the hottest section.  The next two measurement positions at 0.8~m and 1.0~m are similar in temperature.  The third section extends from 1.2 through 2.2~m above the burner and has the shape of a smooth curve decreasing in temperature with increasing elevation.  Comparing the temperatures to the flame height analysis in Chapter 2, the first section aligns with the median flame height, the second section aligns with the portion of the flame that is intermittent, 40~$\%$ of the time or less.  The upper most section of the temperature data aligns with the thermal plume above the flame.

 The second issue is that the FDS predictions do not align equally well with the data from each position, with the best agreement occurring when the burner is positioned against the wall.  Further the temperatures for the burner position, 0.61~m from the wall, was thought to be equivalent to the source fire experiments documented in Chapter 2. However that is not the case.  The installation of the wall resulted in movement of the flame and plume.  The flames oscillate toward the wall and as a result move away from the thermocouples centered over the burner. The resulting decrease in temperature may account for the difference between the plume temperatures and the simulated temperatures from FDS.

@@

Reviewing the videos from a natural gas burner instrumented gypsum wallboard experiment with the burner positioned at 0.310~m from the wall, we can see examples of the flame movement.  In the top row of Figure~\ref{FDS_IWNG_Flame_comp} the three images exhibit points in time where the flame was moving toward the instrumented wall and away from the verticle array of thermocouples that are positioned above the center line of the burner.  The bottom row of images are from the same fire experiment looking at it from the front.  These images also show how the pulsations and movements of the flame change positions relative to the fixed positions of the thermocouples and heat flux gauges.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=1.5in]{../Figures/IWGB_NG_Side1}
  \includegraphics[width=1.5in]{../Figures/IWGB_NG_Side2} 
  \includegraphics[width=1.5in]{../Figures/IWGB_NG_Side3}       \\
 \vspace{5mm}                                                             
  \includegraphics[width=1.5in]{../Figures/IWGB_NG_Front1}   
  \includegraphics[width=1.5in]{../Figures/IWGB_NG_Front2}     
  \includegraphics[width=1.5in]{../Figures/IWGB_NG_Front3}      \\
  \caption[Images of flame movement which are consistant with the measurements and explain the differences between the measured and FDS values.]{FImages of flame movement which are consistant with the measurements and explain the differences between the measured and FDS values.  The natural gas burner was poistioned 0.31~m from the instrumented gypsumboard wall.}
  \label{FDS_IWNG_Flame_comp}
\end{figure}

@@

A flame height range was calculated from each FDS simulation by multiplying the total energy released by 95 and 99~\%, and linking those values to the corresponding height form the *****.
